---
title: "Tidying Data with dplyr and tidyr"
author: "Axel R"
date: "2024-01-20"
output: html_document
---

**Remark**: The material is based on an activity assigned to us by PhD. Renan Escalante.

**Objective:** In this exercise, you will learn how to tidy a dataset using the dplyr and tidyr packages in R. Tidying data is an essential step in data preparation for analysis. We will use a gene expression dataset as an example.

**Prerequisites:** You should have R installed and be familiar with basic data manipulation in R.

**Dataset:** We will use the gene expression dataset provided in the blog post.

# 1. Load the Data

Load the gene expression dataset using the `readr` package and view its dimensions

```{r Load the data, message=FALSE, warning=FALSE}
library(readr)
data <- read_csv("microarray_expression_data.csv", col_names = TRUE)
```

1.  ***What is the purpose of loading the dataset in this step?***

    The `readr` package is a part of the tidyverse in R and is designed for **efficient reading of flat-file data**, such as CSV or TSV files, into R data frames.

2.  ***What does the `dim()` function do? How about `str()`?***

    `dim` function will tell us the dimensions of the dataset we're using. On the other hand, `str()` is used to display the structure of the dataset, providing an overview of the internal structure of the object.

```{r Structure of Data}
dim(data)
str(data)
```

In this case, we have a dataset (`data`) with 5,537 rows and 40 columns. Also, we have several variables in form of character and number.

# 2. Understanding the Untidy Data

Before we start tidying the data, let's understand what makes this dataset untidy. Refer to the blog post to answer the following questions:

1.  ***What are the main reasons that make this dataset untidy?***

    The column called `NAME` has several empty spaces and it has information that doesn't correspond just to the name of the gene, but instead other characteristics that should be encapsulated in other columns.

2.  ***How are the column headers (e.g., `G0.05`, `N0.3`) structured in this dataset?***

    They are structured from lowest to highest concentration (e.g. G0.05, G0.1, G0.15, G0.2 G0.3).

# 3. Separate Multiple Variables in One Column

The `NAME` column contains multiple pieces of information separated by '\|\|'. We need to split this column into separate columns. Perform this operation using the tidyr package's `separate` function.

```{r separate function, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
help("separate")
```

```{r separating data, message=FALSE, warning=FALSE}
library(tidyr)
data_clean <- separate(data = data, col = NAME, into = c("Name", "Function", "Details", "Mutation", "Position"), sep = "\\|\\|", remove = T)
```

1.  ***What are the five new columns created after using the `separate` function?***

    I created the columns `Name`, `Function`, `Details`, `Mutation` and `Position`

2.  ***What does the `sep` argument specify in the `separate` function?***

    How the `NAME` column is gonna be split.

# 4. Remove Whitespace

Some columns have leading and trailing whitespace. Clean these columns

```{r help mutate, eval=FALSE, include=FALSE}
help(mutate)
```

```{r removing whitespaces, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
data_clean <- data_clean %>%
  mutate(Name = str_squish(Name))
```

1.  ***Why is it important to remove leading and trailing whitespace from columns?***

    Because empty spaces can cause errors when running other functions, as well as making the graphs look less clean.

2.  ***How is the `mutate` function used in this step?***

    We are using the **`mutate()`** function to modify a column in the dataframe **`data_clean`**.

    -   **`Name`**: This is the name of the column that will be modified.

    -   **`str_squish(Name)`**: The **`str_squish()`** function from the stringr package is applied to the **`Name`** column. It removes leading, trailing, and excess internal whitespace from character strings.

# 5. Remove Unnecessary Columns

Some columns are not needed for our analysis. Remove columns '`number`', '`GID`', '`YORF`', and '`GWEIGHT`'

```{r Removing columns}
data_clean <- data_clean %>%
  select(c(-Position, -GID, -YORF, -GWEIGHT))
# str(data_clean)
```

1.  ***Why is it important to remove unnecessary columns in data cleaning?***

    It is a crucial step in the data cleaning process as it improves the overall quality, usability, and efficiency of the dataset, making it more suitable for analysis and modeling.

2.  ***How do tidyselect functions work in removing columns?***

    `select()` function is used to select or drop columns from a dataframe. **`c(-Position, -GID, -YORF, -GWEIGHT)`** is a vector specifying the columns to be dropped. The **`-`** sign before each column name indicates that those columns should be removed.

# **6. Reshape Data**

The column headers like `G0.05`, `N0.2`, etc., represent multiple variables.

1.  ***What does the `pivot_longer` function do in this step?***

2.  ***What are the new columns created after using the `pivot_longer` function?***

# 7. **Separate Variables in One Column**

The `sample` column contains two variables: nutrient and rate.

1.  ***What does the `separate` function do in this step?***
2.  ***Why do we use `convert = TRUE` in the `separate` function?***

# 8. **Visualize the Tidied Data**

Visualize the tidied data using `ggplot2` for a gene of interest (e.g., "LEU1") and a biological process of interest (e.g., "leucine biosynthesis").

1.  ***What is the purpose of visualizing data in this step?***

2.  ***How does `ggplot2` help in creating visualizations?***

# 9. **Explore Other Biological Processes**

Explore another biological process (e.g., "sulfur metabolism") using ggplot2 and visualize the gene expression patterns.

1.  ***Why is it important to explore different biological processes in the dataset?***

2.  ***How can you use `facet_wrap` function do in this step?***

# 10. Proffesor's Script

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

```
